# Project

### Author: Qi Meng 


Last modified date: 2018-03-01


========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library('dplyr')
library(ggplot2)
library(tidyr)
library('reshape2')
library(scales)
library(MASS)
library(scales)
library(memisc)
library(lattice)
library(RColorBrewer)
library(ggplot2)
```



```{r echo=FALSE, Load_the_Data}
# Load the Data

pr <- read.csv('~/prosperLoanData.csv')

```


### Loan Data from Prosper

This data set contains 113,937 loans with 81 variables on each loan,
including loan amount, borrower rate (or interest rate), 
current loan status, borrower income, borrower employment status,
borrower credit history, and the latest payment information.



The diamention of Proserper Loan Dataset


```{r}

dim(pr)
```


In total there are 81 variables that corresponding to each loan 
In order better understand the dataset, in this analysis,
19 variables will be selected.



```{r}

pr <- subset(pr, select = c('LoanStatus',
                                'BorrowerAPR',
                                'BorrowerRate',
                                'LenderYield',
                                'ProsperScore',
                                'BorrowerState',
                                'Occupation',
                                'EmploymentStatus',
                                'IsBorrowerHomeowner',
                                'TotalCreditLinespast7years',
                                'TotalInquiries',
                                'BankcardUtilization',
                                'AvailableBankcardCredit',
                                'IncomeRange',
                                'IncomeVerifiable',
                                'LoanOriginalAmount',
                                'LoanOriginationDate',
                                'MonthlyLoanPayment',
                                'Investors'
))
```




```{r}
str(pr)
```




# Univariate Plots Section



Prosper Score measures the loan applicant's risk level, 
the higer the score the lower the risk to lend.


```{r}
summary(pr$ProsperScore)
```

```{r}
table(pr$ProsperScore)
```


The Highest score is 11 and the score distribution would be 
better presented by the following chart.


```{r echo=FALSE ,warning=FALSE, message=FALSE}


ggplot(aes(ProsperScore), data = pr) +
  scale_x_continuous(breaks = seq(1, 11, 1) ) +
  geom_histogram(stat = 'count', binwidth = 1, fill = '#099DD9')  +
    ggtitle(" ProsperScore Histogram ")


```


From the histogram, the majority of the ProserScore are 4, 6, and 8


```{r echo=FALSE ,warning=FALSE, message=FALSE}

ggplot(aes(LoanStatus), data = pr) +
    geom_histogram(stat = 'count', fill = '#099DD9')    + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    ggtitle(" Loan Status Histogram ")

```

Most of the loan are current and the second is the completed loans




```{r}
summary(pr$LoanStatus)
```




```{r echo=FALSE ,warning=FALSE, message=FALSE}

range_list <- c('$0','$1-24,999','$25,000-49,999','$50,000-74,999',
                '$75,000-99,999',
                '$100,000+','Not employed','Not displayed')
pr$income_range <- factor(pr$IncomeRange,levels = range_list)

```



Creating a new variabel income_range to better format the IncomeRange variable




```{r echo=FALSE ,warning=FALSE, message=FALSE}
ggplot(aes(pr$EmploymentStatus), data = pr) +
    geom_histogram(stat = 'count', fill = '#099DD9')    + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggtitle(" EmploymentStatus Histogram ")


```



The loan data mostly coming from people who works employed or full time.



```{r}
summary(pr$EmploymentStatus)
```


```{r echo=FALSE ,warning=FALSE, message=FALSE}

ggplot(aes(pr$income_range), data = pr) +
    geom_histogram(stat = 'count', fill = '#099DD9')    + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggtitle(" Income Range Histogram ")

```


People whose income is between $25,000 and $100,000 applied for the loans.



```{r}
summary(pr$income_range)
```




```{r echo= FALSE, warning= FALSE, message=FALSE}


pr$available_bank_credit_bracket <- cut(pr$AvailableBankcardCredit/1000,
                                        breaks = seq(0, 15, by = 2))
ggplot(aes(available_bank_credit_bracket), 
       data = subset(pr, !is.na(available_bank_credit_bracket)) ) +
    geom_histogram(stat = 'count', fill = '#099DD9') +
    scale_size_area() + 
    labs(size= "Nitrogen",
         x = "Available_bank_credit_bracket(Thousands)",
         y = "Count",
         title = "Available_bank_credit_bracket histogram plot")



```


Based on the available bank credit breakdown, most of the borrowers
have the credit less than 6,000.



```{r}
summary(pr$available_bank_credit_bracket)
```

Summary of available_bank_credit_bracket




```{r echo= FALSE, warning= FALSE, message=FALSE}


# Add borrower_rate_bracket breakdown

pr$borrower_rate_bracket <- cut(pr$BorrowerRate * 100,
                             breaks = seq(0,40,5))


ggplot(aes(borrower_rate_bracket), 
       data = subset(pr, !is.na(borrower_rate_bracket)) ) +
    geom_histogram(stat = 'count', fill = '#099DD9') +
    scale_size_area() + 
    labs(size= "Nitrogen",
         x = "Borrower_Rate (%)",
         y = "Count",
         title = "Borrower_Rate histogram plot")



```

Borrowers' interest rates are between 0 tp 35%.




```{r}
summary(pr$AvailableBankcardCredit)
```


Summary of AvailableBankcardCredit



```{r}
summary(pr$available_bank_credit_bracket)
```


Summary of available_bank_credit_bracket



# Univariate Analysis

### What is the structure of your dataset?

The ProsperLoan data have 113937 overvations and 19 variables
ProsperSocre: A custom risk score built using historical Prosper data.
The score ranges from 1-10, with 10 being the best, or lowest risk score.  
Applicable for loans originated after July 2009.
Loanstatus: Completed, Current, Past Due (1-15 days), Defaulted, Chargedoff,
            Past Due (16-30 days), Cancelled, Past Due (61-90 days), 
            Past Due (31-60 days), Past Due (91-120 days)

EmploymentStatus:Self-employed, Employed, Not available, Full-time,
                Other, Not employed, Part-time, Retired     

IncomeRange: $0, $1-24,999, $25,000-49,999, $50,000-74,999, $75,000-99,999,
            $100,000+, Not employed, Not displayed



### What is/are the main feature(s) of interest in your dataset?
The main feature is the ProsperSocre, which measue the risk ability of the loan itself, verus the BorrowerRate





### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
The loan applicants' occupation,income, bankcard utilization, 
available bank card credit and other variables
might impact the rick score when valued by th Prosper Company



### Did you create any new variables from existing variables in the dataset?
Yes, I created the new variable range_new to reorder
the income range variable in a ascending order.


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy,
adjust, or change the form \
of the data? If so, why did you do this?

Yes. I did select 19 variables out of 81 in total. 
The reason for this is that not all the 
variables are revelent in determining the Prosper Score.






# Bivariate Plots Section


```{r echo=FALSE ,warning=FALSE, message=FALSE}
ggplot(aes(income_range, ProsperScore ), 
       data = subset(pr, !is.na(ProsperScore))) +
  geom_jitter(alpha = .05, shape = 21, colour = '#F79420') +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))  +
    ggtitle(" Income Range and ProsperScore Bivariate Plot ")

```

For applicants with lower income, the Prosper Score seems lower than 6, 
and for people with higher income, 
the ProsperScore seems higher, which means less risk.


```{r}

tapply( pr$ProsperScore, pr$income_range,summary)
```




```{r echo=FALSE ,warning=FALSE, message=FALSE}

ggplot(aes(income_range, ProsperScore ), data = subset(pr, !is.na(ProsperScore))) +
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))  +
    ggtitle(" ProsperScore and Income Range Box Plot ")


```

From the box plot we can see it more clearly, the median of people of 
income greater than $50,000 is much higher 
than people with income less than $50,000




```{r echo=FALSE ,warning=FALSE, message=FALSE}
ggplot(aes(EmploymentStatus, ProsperScore ), data = subset(pr, !is.na(ProsperScore))) +
  geom_jitter(alpha = .05, shape = 21, colour = '#F79420') +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))  +
    ggtitle(" ProsperScore and Employment Status Scatter Plot ")


```


If we compare ProsperScore with employment status, clearly, 
full time employees will have greater ProperScore, and therefore, 
less risky to lend money to them



```{r}
tapply( pr$ProsperScore, pr$EmploymentStatus,summary)
```




```{r echo=FALSE ,warning=FALSE, message=FALSE}
ggplot(aes(EmploymentStatus, ProsperScore ),
       data = subset(pr, !is.na(ProsperScore))) +
    geom_raster( aes(fill = income_range)) + 
    theme(axis.text.x = element_text(angle = 60, hjust = 1))  +
    ggtitle(" ProsperScore and Employment Status Heat Map ")


```


It is clearer to see from the heat map that employed and full time workers who
have above average income will have higher Prosper Score. 
Meanwhile, for some self employed applicants, even though the income range is 
above $100,000, the Prosper Score is extremely low, indicting higher risk than
other applicants with considerably lower income.


```{r echo=FALSE ,warning=FALSE, message=FALSE}
ggplot(aes( ProsperScore, BorrowerRate * 100 ), 
       data = subset(pr, !is.na(ProsperScore))) +
  geom_jitter(alpha = .05, shape = 21, colour = '#F79420') +
  stat_smooth(method = 'lm') +
    ggtitle(" Borrower Rate and Prosper Score Bivariate Plot ")

```


Unsperisingly, applicants with higher prosperscore seem to have lower 
borrow rate.



Summary for Prosper Score and Borrower Rate(%)


```{r}
tapply(pr$BorrowerRate*100, pr$ProsperScore, summary)
```


```{r echo=FALSE ,warning=FALSE, message=FALSE}
with(pr, cor.test(pr$ProsperScore, pr$BorrowerRate))
```


And also from the correlation, Prosper Score and Borrower Rate has a strong 
negative correlation.



```{r echo=FALSE ,warning=FALSE, message=FALSE}
ggplot(aes( EmploymentStatus, BorrowerRate * 100 ), 
       data = subset(pr, !is.na(ProsperScore))) +
  geom_jitter(alpha = .05, shape = 21, colour = '#F79420') +
  stat_smooth(method = 'lm') +
    ggtitle(" Borrower Rate and Employment Status scatter Plot ")

```



```{r echo=FALSE ,warning=FALSE, message=FALSE}

ggplot(aes(pr$AvailableBankcardCredit/1000, pr$BorrowerRate*100 ), 
  data = subset(pr, (!is.na(BorrowerRate)))) +
    
      geom_jitter(alpha =.5, shape = 21, colour = '#099DD9') +
    
      scale_x_continuous(limits = c(1, 150),breaks = seq(1, 150, 20) ) +
      ylim(5,40) +
      ggtitle(" Available Bank Credit and Borrower Rate Scatter Plot ")

```


Higer available bank card credit will also have lower borrow rate. 


```{r}
with(na.omit(pr), cor(BorrowerRate*100,AvailableBankcardCredit/1000))

```


Correlation between BorrowerRate and AvailableBankcardCredit



# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?


Other than main feature, I also noticed applicants with 
higher available bank card credit will also have lower borrow rate. 



### What was the strongest relationship you found?

Borrow rate verus the properscore, The coefficient between them is -0.6682872.





# Multivariate Plots Section


```{r echo=FALSE ,warning=FALSE, message=FALSE}


ggplot(aes( BorrowerRate,ProsperScore), 
       data = subset(pr, !is.na(ProsperScore)) ) +
  geom_jitter(alpha = .05, shape = 21, colour = '#F79420') +
  facet_wrap(~income_range) +
    ggtitle(" Prosper Score and Borrower Rate Scatter Plot ")




```




```{r echo=FALSE ,warning=FALSE, message=FALSE}

ggplot(aes(BorrowerRate,ProsperScore ),
       data = subset(pr, !is.na(ProsperScore))) +
  geom_jitter(alpha = .05, shape = 21, colour = '#099DD9') +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  facet_wrap(~EmploymentStatus)+
    ggtitle(" Prosper Score and Borrower Rate Scatter Plot ")


```


The same pattern also occur for different employment status



```{r echo=FALSE ,warning=FALSE, message=FALSE}

ggplot(aes( ProsperScore,BorrowerRate* 100, colour = income_range), data = pr) + 
  geom_point(alpha = .5, size = .5, position = 'jitter') +
  scale_color_brewer(type =  'div',
                     guide = guide_legend(title = 'income_range', 
                                          reverse = FALSE, 
                                          override.aes = list(alpha = 1,
                                                              size = 2))) +
  scale_y_continuous(limits = c(quantile(pr$BorrowerRate * 100, .05),
                                quantile(pr$BorrowerRate * 100, .90)))+
    ggtitle(" Prosper Score and Borrower Rate Scatter Plot ")



```



```{r echo=FALSE ,warning=FALSE, message=FALSE}

ggplot(aes( ProsperScore,borrower_rate_bracket,
            colour = income_range),
       data = subset(pr, 
                     (!is.na(borrower_rate_bracket)) & 
                         (borrower_rate_bracket != '(35,40]'))) +
    geom_raster( aes(fill = income_range), interpolate = TRUE) +
    ggtitle(" Prosper Score and Borrower Rate Heat Map ")
    
```


When comparing with the main relationship ProsperScore and BorrowerRate, 
we also noticed a liner relationship adding the category 
variable income range



```{r echo=FALSE ,warning=FALSE, message=FALSE}

ggplot(aes(BorrowerRate*100, ProsperScore, colour = EmploymentStatus), 
       data = pr) + 
  geom_point(alpha = .5, size = .75, position = 'jitter') +
  scale_color_brewer(type =  'div', 
                     guide = guide_legend(title = 'EmploymentStatus', 
                                          reverse = FALSE, 
                                          override.aes = list(alpha = 1,
                                                              size = 2))) +
  scale_x_continuous(limits = c(quantile(pr$BorrowerRate* 100, .05),
                                quantile(pr$BorrowerRate*100, .90))) +
    ggtitle(" Prosper Score and Borrower Rate Scatter Plot ")

  
```


```{r echo=FALSE ,warning=FALSE, message=FALSE}

m1 <- lm(I(BorrowerRate) ~ 0 + I(ProsperScore), 
         data = subset(pr, !is.na(ProsperScore)))
m2 <- update(m1, ~ . + AvailableBankcardCredit)
m3 <- update(m2, ~ . + IncomeRange)
m4 <- update(m3, ~ . + EmploymentStatus)
m5 <- update(m4, ~ . + TotalCreditLinespast7years)

mtable(m1, m2, m3, m4, m5)

```



# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

From multivariate analysis, the borrower's rate is also affacted by
other variables such as income range, employment status, 
aviable bank credits.



### Were there any interesting or surprising interactions between features?

Holding the borrower's rate constant,
the employed status will have lower prosper score.



### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.


Yes, I did. For the linear models I created, R square is about 93%,
which means almost 93% percent of the variation can be 
explained by the model. I also excluded the intercept,
which strenthed the linear relationship between the variables. 




------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, warning= FALSE,Plot_One}


ggplot(aes(ProsperScore), data = pr) +
  scale_x_continuous(breaks = seq(1, 11, 1) ) +
  geom_histogram(stat = 'count', binwidth = 1,fill = '#099DD9')  +
    ggtitle(" ProsperScore Histogram ")

```

### Description One

The loans were valued by the risk levels which bing called as ProsperScore,
and the greater the score the lower the risk.
Histogram showed us the counts for different ProsperScore.
Majority of the loans have the score between 4 and 9. 


### Plot Two

```{r echo=FALSE, warning= FALSE, Plot_Two}
ggplot(aes( ProsperScore, BorrowerRate * 100 ),
       data = subset(pr, !is.na(ProsperScore))) +
  geom_jitter(alpha = .05, shape = 21, colour = '#099DD9') +
  stat_smooth(method = 'lm') +
  scale_x_continuous(breaks = seq(0, 10, 1))+
    ggtitle(" Prosper Score and Borrower Rate Scatter Plot ")
```


### Description Two

Ggplot gives us the negative correlation between ProsperScore and 
the borrower's rate, the higher the score seems to 
lead to a lower borrower's rate.




### Plot Three

```{r echo=FALSE, warning= FALSE, Plot_Three}

ggplot(aes( ProsperScore,BorrowerRate* 100, colour = income_range), data = pr) + 
  geom_point(alpha = .5, size = .5, position = 'jitter') +
  scale_color_brewer(type =  'div',
                     guide = guide_legend(title = 'income_range', 
                     reverse = FALSE, 
                     override.aes = list(alpha = 1, size = 2))) +
  scale_y_continuous(limits = c(quantile(pr$BorrowerRate * 100, .05),
                                quantile(pr$BorrowerRate * 100, .90))) +
    ggtitle(" Prosper Score and Borrower Rate Scatter Plot ")

```

### Description Three

When considering other variables, for example, 
income range, will also help us in predicting the borrower's rate.From the plot, 
the higer income leads also have higer ProsperScore,
and revelvantly lower interest rate.


------


# Reflection

Prosper Loan dataset has thorogh loan data regrading theri unique attributes, 
and when evaluating the loan applications, these 
variables could benefit the company in deciding the accurate rate.
From the the analysis, it is shown that borrower's rate the highly 
correlated with borrower's ProsperScore, which measured the risk
of the applicant. 
We also learned that other factors such as income level,
employment status, avaible bank credits could also affect borrower's rate.
More thorough demographic data can be included in the dataset, therefore,
we can better the detailed attributs of the applicants.



